---
title: "Mobile Market Share Animated Donut"
author: "Ansgar Wolsing"
date: "6/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages

```{r packages}
library(tidyverse)
library(gganimate)
library(ggtext)
library(lubridate)
```

## Get the data

Statcounter provides mobile vendor market share back to 2010. The original video goes back to the 1990s. The data can be downloaded in CSV format via https://gs.statcounter.com/vendor-market-share/mobile/worldwide/#monthly-201003-202205
After the download, place the CSV file in your project directory.

```{r read_data}
filename <- "vendor-ww-monthly-201003-202205.csv"
df_raw <- read_csv(filename)

```

## Transform the data

Transform the data to long format for our next steps.

```{r transform_data}
threshold_for_lumping <- 5

df_long <- df_raw %>%
  pivot_longer(cols = -Date, names_to = "vendor", values_to = "market_share") %>% 
  # group vendors with smaller market shares to "Other" based on monthly shares
  mutate(vendor2 = ifelse(market_share < threshold_for_lumping | 
                            vendor == "Unknown", "Other", vendor),
         date = ym(Date)) %>% 
  filter(date > as_date("2010-03-01")) %>% 
  count(date, vendor2, wt = market_share, name = "market_share")

```

The first few rows of the transformed dataframe:

```{r}
head(df_long)
```

All grouped vendors:

```{r}
unique(df_long$vendor2)
```



## First plot with facets

```{r first_plot}

df_long %>% 
  filter(date <= as_date("2011-03-01")) %>% 
  ggplot(aes(vendor2, market_share)) +
  geom_col() +
  coord_flip() +
  facet_wrap(vars(date))

```

## Create a donut chart with ggplot2


{ggpubr} is great for creating several chart types, include donut charts. However, for our animation we will build the plot from scratch in {ggplot2}.


### Start with a pie chart

```{r first_pie}
p <- df_long %>% 
  filter(date == as_date("2022-05-01")) %>%
  arrange(date, market_share) %>% 
  mutate(label_pos = cumsum(market_share) / sum(market_share) 
         - 0.5 * market_share /  sum(market_share),
         label = sprintf("%s<br>%s %%", vendor2, market_share),
         # label = fct_reorder(label, -market_share),
         vendor2 = fct_reorder(vendor2, -market_share)) %>% 
  ggplot(aes(x = 1, market_share, group = vendor2)) +
  geom_col(aes(fill = vendor2), position = "fill") +
  geom_richtext(aes(x = 1.5, label = label, y = label_pos)) +
  paletteer::scale_fill_paletteer_d("palettetown::dratini") +
  coord_polar(theta = "y") +
  guides(fill = "none") +
  theme_void()
p

```


### ... and the donut

We just simply add a white circle on top of the pie chart. Voilà, a donut chart.
Adjust `donut_hole_width` to increase or decrease the donut bar. A value of 0 will result in a pie chart, a value of 1.5 or greater will cover the whole pie chart.

```{r}

donut_hole_width <- 0.9

p_donut <- p + 
  annotate("rect", xmin = 0, xmax = donut_hole_width, ymin = -Inf, ymax = Inf,
           fill = "white") +
  geom_text(aes(x = 0, y = 0, label = format(date, "%B\n%Y")), stat = "unique",
            size = 8)
p_donut

```

## First animation


```{r}

p_donut <- 
  df_long %>% 
  # filter(date == as_date("2022-05-01")) %>%
  # filter(date >= as_date("2020-05-01")) %>%
  mutate(vendor2 = factor(vendor2, levels = unique(df_long$vendor2))) %>% 
  # arrange(date, vendor2) %>% 
  # now we need to calculate the label position within each month
  group_by(date) %>% 
  arrange(desc(vendor2), .by_group = TRUE) %>% 
  mutate(label_pos = cumsum(market_share) / sum(market_share)  
         - 0.5 * market_share / sum(market_share),
         label = sprintf("%s\n%s %%", vendor2, 
                         scales::number(market_share, accuracy = 0.1)),
         label = fct_reorder(label, market_share)
         # vendor2 = fct_reorder(vendor2, -market_share)
         ) %>% 
  ungroup() %>% 
  ggplot(aes(x = 1, market_share, group = vendor2)) +
  geom_col(aes(fill = vendor2), position = "fill") +
  ggrepel::geom_text_repel(aes(x = 1.5, label = label, y = label_pos),
                hjust = 0, family = "Fira Sans",
                min.segment.length = 0, nudge_x = 0.3, point.padding = 1e-05,
                label.padding = 0.3) +
  paletteer::scale_fill_paletteer_d("palettetown::dratini", direction = -1) +
  # paletteer::scale_color_paletteer_d("palettetown::dratini") +
  coord_polar(theta = "y") +
  guides(fill = "none", color = "none") +
  theme_void(base_family = "Fira Sans") + 
  annotate("rect", xmin = 0, xmax = donut_hole_width, ymin = -Inf, ymax = Inf,
           fill = "white") +
  geom_text(aes(x = 0, y = 0, label = format(date, "%B\n%Y")), stat = "unique",
            size = 8)

p_anim <- p_donut +
  transition_states(date)

anim <- animate(p_anim, res = 100, width = 600, height = 600, fps = 10)
anim_save("animated-donut-chart.gif", anim)

```

